---
title: "2_Clean"
author: "Helena"
date: '2023-03-09'
output:
  github_document: default
  word_document:
    reference_docx: word_style1.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Clean
Needless to say, there is quite a bit of cleaning to do before we get to the analysis.

```{r, message = FALSE, warning = FALSE}
#Load the libraries and load the data we assembled in 1_assemble
library(tidyverse)
library(lubridate)

data <- readRDS(here("data", "assembled_data.RDS"))
```

Correct all identified typos. These are things like someone being 660 years old or having their primary treatment 200 years in the future. There will always be typos when data are manually entered into a spreadsheet. Problems like these stand out when doing the exploratory data analysis in 3_CheckQuality. Where I identified issues these were checked against the original patient record and then corrected in the following chunk.

```{r}
data <- data %>%
  mutate(causeofdeath = if_else(rowid == 1958, "Other", causeofdeath),
         dateofpm = if_else(rowid == 2187, "04/04/2016", dateofpm),
         dateofpm = if_else(rowid == 1344, "11/09/2012", dateofpm),
         ageatpm = ifelse(rowid == 1066, 22, ageatpm),
         gender = if_else(rowid == 1066, "M", gender),
         ageatpm = ifelse(rowid == 1369, 57, ageatpm),
         gender = if_else(rowid == 1369, "F", gender),
         ageatpm = ifelse(rowid == 1606, 66, ageatpm),
         uh = ifelse(rowid == 2142, 8.8, uh)) %>%
  filter(rowid != 235)
```

The next sections do various bits of tidying up. First some columns are renamed to make them easier to type!

```{r}
data <- data %>%
  rename(followup = dateoflastfollowupdeath,
         mitcount = mitoticcount40hpf)
```

Next the date columns are converted from character to date class and number columns are made numeric. 

```{r}
data <- data %>%
  mutate(dateofpm = dmy(dateofpm),
         followup = dmy(followup),
         ageatpm = as.numeric(ageatpm),
         lbd = as.numeric(lbd),
         uh = as.numeric(uh),
         mitcount = as.numeric(mitcount),
         onsetofmm = dmy(onsetofmm))
```

Also there are some quirks specific to this dataset that needed correction - they are slightly difficult to explain, but feel free to ask if you need more info. The dataset used death notifications from the national cancer registry as one source of follow up data and it was thought that the last update from the registry was received in "2020-05-01" (so if people were not listed they were assumed alive on that date), but when this was checked the real date was "2018-09-01". Therefore all people listed as "Alive" with last follow up on that date, needed the date changed. Also some people did not have any follow up listed (normally overseas patients) which may make it appear as though their follow-up occurred before their primary treatment (normally the day before). In these cases I set follow up to equal the day of their primary treatment.

```{r}
data <- data %>%
  mutate(followup = if_else(followup == as_date("2020-05-01") & status != "D", as_date("2018-09-01"), followup),
         followup = if_else(followup < dateofpm, dateofpm, followup))
```

I also made some new columns I needed for data exploration or analysis. I calculated length of follow up and time to detection of mets in either years or months (using my favourite operator %--%). I also extracted the year of primary treatment into a separate column.

```{r}
data <- data %>%
  mutate(lfumonths = dateofpm %--% followup / dmonths(1),
         lfuyears = dateofpm %--% followup /dyears(1),
         lfumets = dateofpm %--% onsetofmm / dyears(1),
         dateofpmyear = year(dateofpm))
```

Next, the factor fields! So the primary treatment and cause of death fields have just been manually enetered as free text and therefore contain some inconsistencies. Here I combine bunches of things together so we are just left with the categories which are useful for analysis.

```{r}
data <- data %>%
  mutate(primarytreatment = str_to_lower(primarytreatment),
         primarytreatment = str_replace_all(primarytreatment, " ", ""),
         primarytreatment = fct_collapse(primarytreatment, `endoresection+prxt` = c("endoresectionandprxt",
                                                                                    "endoresectionandplaque",
                                                                                    "endoresection+prxt",
                                                                                    "endoresection+plaque",
                                                                                    "endoresection")),
         primarytreatment = fct_collapse(primarytreatment, `localresection+prxt` = c("localresection",
                                                                                     "localresection+prxt")),
         secondary = if_else(is.na(secondarytreatment), "None", "Secondary Treatment")) %>%
  mutate(causeofdeath = if_else(status == "D" & is.na(causeofdeath), "Unknown", causeofdeath),
         causeofdeath = as.factor(causeofdeath),
         causeofdeath = fct_collapse(causeofdeath, MM = c("MM", "Brain mets"),
                                     Other = c("Other", "Myocardial infarction")),
         causeofdeath = factor(causeofdeath, levels = c("MM", "Other", "Unknown")))
```

Lastly, tidying up the genetic columns. For our purposes chr3 should either be "N" (normal), "L" (loss) or NA and chr8q should either be "G" (gain), "N" (normal) or NA. This sorts that out.

```{r}
data <- data %>%
  mutate(chr3 = na_if(chr3, "U"),
         chr3 = if_else(chr3 == "G", "N", chr3),
         chr8q = na_if(chr8q, "U"),
         chr8q = if_else(chr8q == "L", "N", chr8q),
         with3 = if_else(is.na(chr3), "No result", "Chr3 result"))
```

Finally, apply some exclusion criteria. The analysis only includes people having their primary treatment 2007-2016, those people who have the minimum data required for the LUMPOIII website. It excludes people who already had detectable mets at primary diagnosis and those who had non-standard primary treatments.

```{r}
# Save the dataset
saveRDS(data, here("data", "processed_data.RDS"))
```

The dataset should now look something like this (the following data is made up):

```{r, echo=FALSE}
fake_pd <- readRDS(here("data", "fake_processed_data.RDS"))
glimpse(fake_pd)
```

