---
title: "Analysis"
author: "Helena"
date: '2023-02-27'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose of this document

This is intended to explain the analytic code for the paper 'Sensitivity and specificity of different prognostic systems when used to stratify for surveillance in a population of uveal melanoma patients'. I would recommend reading the paper before trying to understand the code. There is an R script called 'Analysis' in the 'scripts' folder. Running that script produces all the plots and summaries that are in the paper. This document walks through the code in that script.

This code loads all the required packages and the dataset:
```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(plotROC)
library(ROCit)
library(scales)
library(here)
library(PropCIs)
source(here("scripts", "Functions.R"))

roc_data <- readRDS(here("data", "roc_data.RDS"))
```

At this point the dataset has been prepared for analysis. The preparation steps will be addressed in a separate markdown file. The data should now look like this:

```{r}
str(roc_data)
```


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
