---
title: "Analysis"
author: "Helena"
date: '2023-02-27'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Purpose of this document

This is intended to explain the analytic code for the paper 'Sensitivity and specificity of different prognostic systems when used to stratify for surveillance in a population of uveal melanoma patients'. I would recommend reading the paper before trying to understand the code. There is an R script called 'Analysis' in the 'scripts' folder. Running that script produces all the plots and summaries that are in the paper. This document walks through the code in that script. The data is pseudonimised but not publicly available because it is sensitive data and there is a danger of re-identification. 

This code loads all the required packages, custom functions and the dataset:
```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(plotROC)
library(ROCit)
library(patchwork)
library(scales)
library(here)
library(PropCIs)
source(here("scripts", "Functions.R"))

roc_data <- readRDS(here("data", "roc_data.RDS"))
```

```{r, echo = FALSE}
fake_roc_data <- readRDS(here("data", "fake_roc_data.RDS"))
```


At this point the dataset has been prepared for analysis. The preparation steps will be addressed in a separate markdown file. The data should now look like this (the following example data is made up):

```{r}
glimpse(fake_roc_data)
```

NOTE: cols 'MM' is the LUMPO 5 year MAM and 'score' is the LPM 5 year MAM. I agree I could've called them something more helpful.

One last preparation step - the risk score output from LPM is from 0-100 rather than between 0-1. It needs to be the latter for the ROC analysis plots and functions that come later. This is why it gets divided by 100.
```{r}
roc_data <- roc_data %>%
  mutate(
    score = score / 100,
  )
```


### Table 2: Descriptive statistics

This is all self-explanatory so I've just included a few example lines and all the outputs went into Table 2 of the paper.I used either count(), median(), or range() for each column as appropriate. Note that all the counts, medians and ranges are all consistent with previously published estimates so don't raise any data quality concerns. 
```{r}
roc_data %>%
  count(outcome)

median(roc_data$ageatpm)
range(roc_data$ageatpm)
```

### Figure 1: Distribution plots

These plots show the distribution of risk scores / categories in the dataset under the four different prognostic systems with red dashed lines showing example cut-offs which could be used to stratify into high and low risk for enrollment in surveillance. Only LUMPO and AJCC are shown in the paper due to space.

```{r}
lumpo_dist <- roc_data %>%
  ggplot(aes(MM)) +
  stat_bin(binwidth = 0.05, boundary = 0, color = "black", fill = "black", alpha = 0.4) +
  labs(
    x = "LUMPO 5-year MAM (binwidth = 0.05)",
    title = "A: LUMPO III"
  ) +
  theme_classic()

lumpo_dist

ajcc_dist <- roc_data %>%
  mutate(stage = factor(stage, levels = c("I", "IIA", "IIB", "IIIA", "IIIB", "IIIC"))) %>%
  ggplot(aes(stage)) +
  geom_bar(color = "purple", fill = "purple", alpha = 0.5) +
  labs(title = "B: AJCC stage") +
  theme_classic()

ajcc_dist

roc_data %>%
  ggplot(aes(score)) +
  stat_bin(binwidth = 0.1, boundary = 0, color = "darkgoldenrod", fill = "#E69F00") +
  labs(
    x = "LPM 5-year MAM (binwidth = 0.1)",
    title = "C: LPM"
  ) +
  theme_classic()

bar_chr3 <- roc_data %>%
  mutate(
    chr3 = if_else(chr3 == "L", "Monosomy", "Disomy"),
    chr3 = if_else(is.na(chr3), "Unknown", chr3)
  ) %>%
  ggplot(aes(chr3)) +
  geom_bar(color = "pink3", fill = "pink") +
  labs(x = "Chromosome 3 status", title = "D: Genetics") +
  theme_classic()

bar_chr3
```

### Table 3: Stratify the dataset into high and low risk categories

Stratifying the dataset means deciding on a cutoff value above which patients are classified as 'high-risk' and therefore enrolled in surveillance. In the example below, all patients to the right of the red dashed line are considered high-risk.

```{r, echo = FALSE}
lumpo_dist + 
  geom_vline(xintercept = 0.1, color = "red", linetype = "dashed")
```

The following chunk creates a table of sensitivity and specificity estimates etc at different LUMPO cut-off values. In the script the same table gets produced for LPM. These tables produce the figures for Table 3 in the paper. You can look at different cut-offs by changing the figures included in the 'cut-off' column). calc_prevalence and ci_prev are custom functions in the Functions.R script.

```{r}
prevalence <- calc_prevalence(roc_data)
ci_prev <- calc_ci_prev(roc_data)

cuts_tibble_mm <- tibble(cutoff = c(0.5, 0.3, 0.2, 0.1, 0.05)) %>% # LUMPO
  rowwise() %>%
  mutate(
    sensitivity = calc_sens1(roc_data, MM, cutoff),
    specificity = calc_spec1(roc_data, MM, cutoff),
    TP = prevalence * sensitivity,
    TN = (1 - prevalence) * specificity,
    FN = prevalence - TP,
    FP = (1 - prevalence) - TN,
    NPV = TN / (TN + FN),
    PPV = TP / (FP + TP),
    surveillance = TP + FP,
    sens_lower = calc_sens_ci1(roc_data, MM, cutoff)[[1]][1],
    sens_upper = calc_sens_ci1(roc_data, MM, cutoff)[[1]][2],
    spec_lower = calc_spec_ci1(roc_data, MM, cutoff)[[1]][1],
    spec_upper = calc_spec_ci1(roc_data, MM, cutoff)[[1]][2],
    inv_spec = 1 - specificity,
    inv_spec_upper = 1 - spec_upper,
    inv_spec_lower = 1 - spec_lower
  )

cuts_tibble_mm
```
```{r, include = FALSE}
cuts_tibble_par <- tibble(cutoff = c(0.3, 0.2, 0.1, 0.05, 0.04, 0.03)) %>% # LPM
  rowwise() %>%
  mutate(
    sensitivity = calc_sens1(roc_data, score, cutoff),
    specificity = calc_spec1(roc_data, score, cutoff),
    TP = prevalence * sensitivity,
    TN = (1 - prevalence) * specificity,
    FN = prevalence - TP,
    FP = (1 - prevalence) - TN,
    NPV = TN / (TN + FN),
    PPV = TP / (FP + TP),
    surveillance = TP + FP,
    sens_lower = calc_sens_ci1(roc_data, score, cutoff)[[1]][1],
    sens_upper = calc_sens_ci1(roc_data, score, cutoff)[[1]][2],
    spec_lower = calc_spec_ci1(roc_data, score, cutoff)[[1]][1],
    spec_upper = calc_spec_ci1(roc_data, score, cutoff)[[1]][2],
    inv_spec = 1 - specificity,
    inv_spec_upper = 1 - spec_upper,
    inv_spec_lower = 1 - spec_lower
  )
```

I did this a bit differently for AJCC because it's split into categories rather than a continuous score. For each of the cut-offs (red dashed lines) displayed below I've added a column to the dataset with the categorisation for each patient under the system, then made a summary table for that column. (Yes, this is a bit long winded and there is probably a better way)

```{r, echo = FALSE}
ajcc_dist +
  geom_vline(xintercept = 1.5, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2.5, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 3.5, color = "red", linetype = "dashed")
  
```

```{r}
roc_data <- roc_data %>%
  mutate(
    stage_2a = if_else(stage %in% c("I"), "No surveillance", "Surveillance"),
    stage_2b = if_else(stage %in% c("I", "IIA"), "No surveillance", "Surveillance"),
    stage_3a = if_else(stage %in% c("I", "IIA", "IIB"), "No surveillance", "Surveillance")
  )

stage2a_tibble <- tibble(
  sensitivity = calc_sens2(roc_data, stage_2a),
  specificity = calc_spec2(roc_data, stage_2a),
  TP = prevalence * sensitivity,
  TN = (1 - prevalence) * specificity,
  FN = prevalence - TP,
  FP = (1 - prevalence) - TN,
  NPV = TN / (TN + FN),
  PPV = TP / (FP + TP),
  surveillance = TP + FP,
  sens_lower = calc_sens_ci2(roc_data, stage_2a)[[1]][1],
  sens_upper = calc_sens_ci2(roc_data, stage_2a)[[1]][2],
  spec_lower = calc_spec_ci2(roc_data, stage_2a)[[1]][1],
  spec_upper = calc_spec_ci2(roc_data, stage_2a)[[1]][2],
  inv_spec = 1 - specificity,
  inv_spec_upper = 1 - spec_upper,
  inv_spec_lower = 1 - spec_lower
  )

stage2a_tibble
```
```{r, include = FALSE}
stage2b_tibble <- tibble(
  sensitivity = calc_sens2(roc_data, stage_2b),
  specificity = calc_spec2(roc_data, stage_2b),
  TP = prevalence * sensitivity,
  TN = (1 - prevalence) * specificity,
  FN = prevalence - TP,
  FP = (1 - prevalence) - TN,
  NPV = TN / (TN + FN),
  PPV = TP / (FP + TP),
  surveillance = TP + FP,
  sens_lower = calc_sens_ci2(roc_data, stage_2b)[[1]][1],
  sens_upper = calc_sens_ci2(roc_data, stage_2b)[[1]][2],
  spec_lower = calc_spec_ci2(roc_data, stage_2b)[[1]][1],
  spec_upper = calc_spec_ci2(roc_data, stage_2b)[[1]][2],
  inv_spec = 1 - specificity,
  inv_spec_upper = 1 - spec_upper,
  inv_spec_lower = 1 - spec_lower
  )
stage3a_tibble <- tibble(
  sensitivity = calc_sens2(roc_data, stage_3a),
  specificity = calc_spec2(roc_data, stage_3a),
  TP = prevalence * sensitivity,
  TN = (1 - prevalence) * specificity,
  FN = prevalence - TP,
  FP = (1 - prevalence) - TN,
  NPV = TN / (TN + FN),
  PPV = TP / (FP + TP),
  surveillance = TP + FP,
  sens_lower = calc_sens_ci2(roc_data, stage_3a)[[1]][1],
  sens_upper = calc_sens_ci2(roc_data, stage_3a)[[1]][2],
  spec_lower = calc_spec_ci2(roc_data, stage_3a)[[1]][1],
  spec_upper = calc_spec_ci2(roc_data, stage_3a)[[1]][2],
  inv_spec = 1 - specificity,
  inv_spec_upper = 1 - spec_upper,
  inv_spec_lower = 1 - spec_lower
  )
```

Lastly, for the monosomy 3 system I again add a column (m3_str) to the dataset classifying each patient. If a patient has monosomy 3 ('L') or no chromosome 3 result they are placed in the 'Surveillance' category. This method of classifying was based on a description of current practice at LOOC. Again this is summarised in a table.
```{r}
bar_chr3 +
  geom_vline(xintercept = 1.5, color = "red", linetype = "dashed")
```

```{r}
roc_data <- roc_data %>%
  mutate(
    m3_str = if_else(chr3 == "L" | is.na(chr3), "Surveillance", "No surveillance")
  )

m3_tibble <- tibble(
  sensitivity = calc_sens2(roc_data, m3_str),
  specificity = calc_spec2(roc_data, m3_str),
  TP = prevalence * sensitivity,
  TN = (1 - prevalence) * specificity,
  FN = prevalence - TP,
  FP = (1 - prevalence) - TN,
  NPV = TN / (TN + FN),
  PPV = TP / (FP + TP),
  surveillance = TP + FP,
  sens_lower = calc_sens_ci2(roc_data, m3_str)[[1]][1],
  sens_upper = calc_sens_ci2(roc_data, m3_str)[[1]][2],
  spec_lower = calc_spec_ci2(roc_data, m3_str)[[1]][1],
  spec_upper = calc_spec_ci2(roc_data, m3_str)[[1]][2]
) %>%
  mutate(
    inv_spec = 1 - specificity,
    inv_spec_upper = 1 - spec_upper,
    inv_spec_lower = 1 - spec_lower
  )
```

### ROC Analysis: Calculate AUCs with cis
The AUCs relate to the ROC curves in figure 2 and are given in the text of the paper.
```{r}
roc_lumpo <- rocit(
  score = roc_data$MM,
  class = roc_data$outcome, negref = "No mets"
)
paste("roc_data lumpo", auc_label(roc_lumpo))
```

### Figure 2
To make the plots that have the roc curves as well as labelled point estimates I've used geom_roc, then overlaid points, labels and confidence intervals as annotations because ggplot2 doesn't support the combination of points and roc curves (which is why the code for it is so long!). If you can show me a better way I will buy you a coffee.
```{r}
longdata <- melt_roc(roc_data, "outcome_binary", c("MM", "score")) %>%
  mutate(
    name = if_else(name == "MM", "LUMPO III", "LPM"),
    name = factor(name, levels = c("LUMPO III", "LPM"))
  )

fig2 <- ggplot(longdata, aes(d = D, m = M, color = name)) +
  scale_color_manual(values = c("black", "#E69F00")) +
  annotate(
    geom = "rect", xmin = m3_tibble$inv_spec_upper, xmax = m3_tibble$inv_spec_lower, ymin = m3_tibble$sens_lower, ymax = m3_tibble$sens_upper,
    fill = "pink", alpha = 0.5
  ) +
  annotate(
    geom = "rect", xmin = stage2a_tibble$inv_spec_upper, xmax = stage2a_tibble$inv_spec_lower, ymin = stage2a_tibble$sens_lower, ymax = stage2a_tibble$sens_upper,
    fill = "purple", alpha = 0.5
  ) +
  annotate(
    geom = "rect", xmin = stage2b_tibble$inv_spec_upper, xmax = stage2b_tibble$inv_spec_lower, ymin = stage2b_tibble$sens_lower, ymax = stage2b_tibble$sens_upper,
    fill = "purple", alpha = 0.5
  ) +
  annotate(
    geom = "rect", xmin = stage3a_tibble$inv_spec_upper, xmax = stage3a_tibble$inv_spec_lower, ymin = stage3a_tibble$sens_lower, ymax = stage3a_tibble$sens_upper,
    fill = "purple", alpha = 0.5
  ) +
  annotate(
    geom = "rect", xmin = cuts_tibble_mm$inv_spec_upper[1], xmax = cuts_tibble_mm$inv_spec_lower[1],
    ymin = cuts_tibble_mm$sens_lower[1], ymax = cuts_tibble_mm$sens_upper[1],
    fill = "black", alpha = 0.4
  ) +
  annotate(
    geom = "rect", xmin = cuts_tibble_mm$inv_spec_upper[3], xmax = cuts_tibble_mm$inv_spec_lower[3],
    ymin = cuts_tibble_mm$sens_lower[3], ymax = cuts_tibble_mm$sens_upper[3],
    fill = "black", alpha = 0.4
  ) +
  annotate(
    geom = "rect", xmin = cuts_tibble_mm$inv_spec_upper[4], xmax = cuts_tibble_mm$inv_spec_lower[4],
    ymin = cuts_tibble_mm$sens_lower[4], ymax = cuts_tibble_mm$sens_upper[4],
    fill = "black", alpha = 0.4
  ) +
  annotate(
    geom = "rect", xmin = cuts_tibble_mm$inv_spec_upper[5], xmax = cuts_tibble_mm$inv_spec_lower[5],
    ymin = cuts_tibble_mm$sens_lower[5], ymax = cuts_tibble_mm$sens_upper[5],
    fill = "black", alpha = 0.4
  ) +
  geom_roc(n.cuts = 0, labels = FALSE) +
  labs(x = "1-specificity", y = "sensitivity") +
  geom_point(x = 1 - m3_tibble$specificity, y = m3_tibble$sensitivity, color = "#F8766D") +
  geom_point(x = stage2a_tibble$inv_spec, y = stage2a_tibble$sensitivity, color = "purple") +
  geom_point(x = stage2b_tibble$inv_spec, y = stage2b_tibble$sensitivity, color = "purple") +
  geom_point(x = stage3a_tibble$inv_spec, y = stage3a_tibble$sensitivity, color = "purple") +
  geom_point(x = cuts_tibble_mm$inv_spec[1], y = cuts_tibble_mm$sensitivity[1], color = "black") +
  annotate("text",
    x = cuts_tibble_mm$inv_spec[1] - 0.05, y = cuts_tibble_mm$sensitivity[1],
    label = cuts_tibble_mm$cutoff[1], size = 4
  ) +
  geom_point(x = cuts_tibble_mm$inv_spec[3], y = cuts_tibble_mm$sensitivity[3], color = "black") +
  annotate("text",
    x = cuts_tibble_mm$inv_spec[3] - 0.06, y = cuts_tibble_mm$sensitivity[3],
    label = cuts_tibble_mm$cutoff[3], size = 4
  ) +
  geom_point(x = cuts_tibble_mm$inv_spec[4], y = cuts_tibble_mm$sensitivity[4], color = "black") +
  annotate("text",
    x = cuts_tibble_mm$inv_spec[4], y = cuts_tibble_mm$sensitivity[4] + 0.06,
    label = cuts_tibble_mm$cutoff[4], size = 4
  ) +
  geom_point(x = cuts_tibble_mm$inv_spec[5], y = cuts_tibble_mm$sensitivity[5], color = "black") +
  annotate("text",
    x = cuts_tibble_mm$inv_spec[5], y = cuts_tibble_mm$sensitivity[5] + 0.05,
    label = cuts_tibble_mm$cutoff[5], size = 4
  ) +
  annotate("text",
    x = stage2a_tibble$inv_spec + 0.13, y = stage2a_tibble$sensitivity, label = ">= IIA", size = 4,
    color = "mediumorchid4"
  ) +
  annotate("text",
    x = stage2b_tibble$inv_spec + 0.13, y = stage2b_tibble$sensitivity, label = ">= IIB", size = 4,
    color = "mediumorchid4"
  ) +
  annotate("text",
    x = stage3a_tibble$inv_spec + 0.13, y = stage3a_tibble$sensitivity, label = ">= IIIA", size = 4,
    color = "mediumorchid4"
  ) +
  geom_point(x = 0.68, y = 0.08, color = "#F8766D") +
  annotate("text", x = 0.9, y = 0.08, label = "Monosomy 3", size = 3) +
  geom_point(x = 0.68, y = 0.03, color = "purple") +
  annotate("text", x = 0.9, y = 0.03, label = "AJCC system", size = 3) +
  theme_classic() +
  theme(legend.position = c(0.8, 0.25)) +
  theme(legend.title = element_blank()) +
  theme(legend.background = element_blank())

fig2
```

### Splitting the data into two different subpopulations
Chromosome 3 is an important prognostic factor in LPM and LUMPO and a non-random subset of patients do not have a chromosome 3 result (discussed in paper). Therefore I wanted to consider the performance of the models on each subpopulation separately. This code splits the dataset into these two subpopulations.
```{r}
roc_data <- roc_data %>%
  mutate(
    with3 = as_factor(if_else(is.na(chr3), "No result", "Chr3 result")),
    area = lbd * uh,
  )

roc_data2 <- roc_data %>%
  filter(with3 == "Chr3 result")

roc_data3 <- roc_data %>%
  filter(with3 != "Chr3 result")
```
The differences between these two subpopulations are summarised in figure s3
```{r}
prevalence2 <- calc_prevalence(roc_data2)
ci_prev2 <- calc_ci_prev(roc_data2)

prevalence3 <- calc_prevalence(roc_data3)
ci_prev3 <- calc_ci_prev(roc_data3)

prev_tibble <- tibble(
  subpopulation = c("Chr3 result", "No result"),
  prevalence = c(prevalence2, prevalence3),
  prev_low = c(ci_prev2[[1]][[1]], ci_prev3[[1]][[1]]),
  prev_high = c(ci_prev2[[1]][[2]], ci_prev3[[1]][[2]])
)

figs3a <- prev_tibble %>% ggplot(aes(x = subpopulation, y = prevalence, ymax = prev_high, ymin = prev_low)) +
  geom_bar(stat = "identity") +
  geom_errorbar(width = 0.1) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_classic() +
  labs(title = "A: Prevalence of outcome")

figs3b <- roc_data %>%
  ggplot(aes(with3, area)) +
  geom_boxplot() +
  labs(x = NULL, y = "tumour size (mm^2)", title = "B: Tumour size in subpopulations") +
  theme_classic()

figs3a + figs3b
```

All the code discussed previously is repeated for roc_data2 and roc_data3 to result in figure 3.

### Figure 4: LUMPO strategy with different cut-offs for two subpopulations
Given that LUMPO and LPM are less accurate without a chromosome 3 result, it was decided that a more conservative cut-off should be used so that high sensitvity is maintained for patients lacking chromosome 3. The following code applies a different cut-off depending on whether a chromosome 3 result is present, then summarises that sensitvty and specificity of this strategy overall.
```{r}
roc_data <- roc_data %>% mutate(
  fig4_str = case_when(is.na(chr3) & MM >= 0.045 ~ "Surveillance",
                       is.na(chr3) & MM < 0.045 ~ "No surveillance",
                       !is.na(chr3) & MM >= 0.07 ~ "Surveillance",
                       !is.na(chr3) & MM < 0.07 ~ "No surveillance")
)

fig4_lumpo <- tibble(sensitivity = calc_sens2(roc_data, fig4_str),
                     specificity = calc_spec2(roc_data, fig4_str),
                     TP = prevalence * sensitivity,
                     TN = (1 - prevalence) * specificity,
                     FN = prevalence - TP,
                     FP = (1 - prevalence) - TN,
                     NPV = TN / (TN + FN),
                     PPV = TP / (FP + TP),
                     surveillance = TP + FP,
                     sens_lower = calc_sens_ci2(roc_data, fig4_str)[[1]][1],
                     sens_upper = calc_sens_ci2(roc_data, fig4_str)[[1]][2],
                     spec_lower = calc_spec_ci2(roc_data, fig4_str)[[1]][1],
                     spec_upper = calc_spec_ci2(roc_data, fig4_str)[[1]][2]) %>%
  mutate(inv_spec = 1 - specificity,
         inv_spec_upper = 1 - spec_upper,
         inv_spec_lower = 1 - spec_lower)

# Plot 3 strategies
strategies <- tibble(
  strategy = factor(c("LUMPO III" ,"Monosomy 3", "AJCC"), levels = c("LUMPO III" ,"Monosomy 3", "AJCC")),
  sensitivity = c(fig4_lumpo$sensitivity, m3_tibble$sensitivity, stage2a_tibble$sensitivity),
  specificity = c(fig4_lumpo$specificity, m3_tibble$specificity, stage2a_tibble$specificity),
  sens_upper = c(fig4_lumpo$sens_upper, m3_tibble$sens_upper, stage2a_tibble$sens_upper),
  sens_lower = c(fig4_lumpo$sens_lower, m3_tibble$sens_lower, stage2a_tibble$sens_lower),
  spec_upper = c(fig4_lumpo$spec_upper, m3_tibble$spec_upper, stage2a_tibble$spec_upper),
  spec_lower = c(fig4_lumpo$spec_lower, m3_tibble$spec_lower, stage2a_tibble$spec_lower)
) %>%
  mutate(
    inv_spec = 1 - specificity,
    inv_spec_upper = 1 - spec_upper,
    inv_spec_lower = 1 - spec_lower
  )

strategies %>% ggplot(aes(x = inv_spec, y = sensitivity)) +
  geom_hline(aes(yintercept = 0.95), linetype = "dashed", color = "grey", alpha = 0.5) +
  geom_rect(aes(xmin = inv_spec_upper, xmax = inv_spec_lower, ymin = sens_lower,
                ymax = sens_upper, fill = strategy), alpha = 0.5) +
  geom_point(aes(color = strategy)) +
  geom_abline(aes(intercept = 0, slope = 1), color = "white") +
  scale_x_continuous("1-specificity", limits = c(0, 1)) +
  scale_y_continuous("sensitivity", limits = c(0, 1)) +
  scale_color_manual(values = c("black", "#F8766D", "purple")) +
  scale_fill_manual(values = c("black", "pink", "purple")) +
  theme(legend.position = (c(0.85, 0.25))) +
  labs(
    title = "B: Strategy summary" ) +
  theme_classic() +
  theme(legend.position = c(0.8, 0.25)) +
  theme(legend.title = element_blank()) +
  theme(legend.background = element_blank())
```

